Fully Extrinsic Type Inference (DRAFT 2025-06-20)

### 2025-08-08 doc/extrinsic_type_inference_paper/paper.tex
update notation for factoring column from lfp

\newcommand{\factorsinto}{\Vdash}

@@

\newcommand{\multi}[1]{\widebar{\ #1\ }}
\newcommand{\hastype}{:}
\newcommand{\pattype}{\ \lozenge\ }
\newcommand{\restricted}{\ \blacklozenge\ }
\newcommand{\liftfun}{:}

@@

    \inferrule {
      \TODO{modify with case for non-recursive parameter type}
      \\\\
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
      \\
      \alpha \notin \textbf{ftv}(\Delta')
      \\\\
      \alpha \downarrow \Pi
      \fallingdotseq 
      \alpha' \uparrow \Pi'
      \\
      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
      \\\\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{left} \mapsto \tau_l
      \\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{right} \mapsto \tau_r
    } {
      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
      \hastype 
      \tau_l \J{->} \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \entails \Omega \given \Theta', \Delta'
      \\
      \Theta' \cup \Theta_l, \Delta' \entails
      \tau_l \subtypes \tau_r \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails
      \J{EXI[}\Theta_l\J{]}\Omega \J{:} \tau_l \subtypes \tau_r
      \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Theta, \Delta \Vvdash \Omega
      \\
      \Theta, \Delta \given \Omega \entails \Theta', \Delta'
      \\
      \Theta' \cup \Theta_r, \Delta' \entails \tau_l \subtypes \tau_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails \tau_l \subtypes \J{ALL[}\Theta_r\J{]}\Omega \J{:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau_l \subtypes l \mapsto \tau_f 
      \\
      \Theta, \Delta \entails \tau_f \subtypes \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \tau_l \typdiff \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{Update proof, deriving that lower does not intersect with subtraction}
      \\\\
      \TODO{in lean, prove that subtyping is equivalent to decidable definition under restricted subtyping form}
      \\\\
      \epsilon \pattype \tau_r
      \\
      \Theta, \Delta \restricted \tau_r \subtypes  \tau
      \\\\
      \Theta, \Delta \not\Vvdash \tau \subtypes \tau_r
      \\
      \Theta, \Delta \not\Vvdash \tau_r \subtypes \tau
      \\
      \Theta, \Delta \entails \tau \subtypes \tau_l \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \tau_l \typdiff \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\noindent
For least fixed point introduction \ms{
\tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r
}, the solver has a couple options.
One option is to check for a syntactic guarantee that the least fixed point is inflatable without diverging 
\ms{
... \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
} and then solve the inflated introduction.
Alternatively, it can strengthen the least fixed point
with union drop \ms{\alpha \Vdash \tau_r \supertypes \tau_r'},
which drops any parts of the union containing the self reference.

@@

\noindent
Refinement elimination is the dual of expansion introduction, so its rules
rely on analogous techniques.
However, there is one additional rule
for eliminating an intersection and introducing an implication \ms{
\tau_l \& \tau_r \subtypes \tau_a \J{->} \tau_b
}.
As we mentioned earlier when discussing packing together the inferred types of 
multiple paths of functions, sometimes the argument type is weaker than each
path type antecedent, yet could be stronger than union of all the path type antecedents.    
As such, the solver may leverage function type factorization \ms{
\Vdash \tau_l \J{\&} \tau_r \subtypes \tau_x \rightarrow \tau_y 
} to weaken a function type
into a single path type where the antecedent and the consequent may consist
of compounded unions.

@@

\begin{requirement}
  \label{req:factorization_soundness}
  \emph{factorization soundness}  
  \small
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau \subtypes l \mapsto \tau'
    } {
      \forall e' \qua
      e' \hastype \tau' 
      \iff 
      (\exists e \qua e \J{.} l \cong e' \up \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau)
    }
  \end{mathpar}
\end{requirement}

>>>>

@@

\newcommand{\multi}[1]{\widebar{\ #1\ }}
\newcommand{\hastype}{:}
\newcommand{\pattype}{\ \lozenge\ }
\newcommand{\restricted}{\ \blacklozenge\ }
\newcommand{\dentails}{\Vdash}
\newcommand{\liftfun}{:}

@@

    \inferrule {
      \TODO{modify with case for non-recursive parameter type}
      \\\\
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
      \\
      \alpha \notin \textbf{ftv}(\Delta')
      \\\\
      \alpha \downarrow \Pi
      \fallingdotseq 
      \alpha' \uparrow \Pi'
      \\
      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
      \\\\
      \textbf{factor}(\alpha' \entails \tau' \cdot \J{left}) = \tau_l
      \\
      \textbf{factor}(\alpha' \entails \tau' \cdot \J{right}) = \tau_r
    } {
      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
      \hastype 
      \tau_l \J{->} \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Theta, \Delta \restricted \Omega
      \\
      \Theta, \Delta \entails \Omega \given \Theta', \Delta'
      \\
      \Theta' \cup \Theta_l, \Delta' \entails
      \tau_l \subtypes \tau_r \given \Theta'', \Delta''
    } {
      \Theta, \Delta \entails
      \J{EXI[}\Theta_l\J{]}\Omega \J{:} \tau_l \subtypes \tau_r
      \given \Theta'', \Delta'' 
    }
  \end{mathpar}
\end{definition}

@@

    \inferrule {
      \Theta, \Delta \restricted \Omega
      \\
      \Theta, \Delta \given \Omega \entails \Theta', \Delta'
      \\
      \Theta' \cup \Theta_r, \Delta' \entails \tau_l \subtypes \tau_r \given \Theta'', \Delta'' 
    } {
      \Theta, \Delta \entails \tau_l \subtypes \J{ALL[}\Theta_r\J{]}\Omega \J{:} \tau_r \given \Theta'', \Delta''
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_lfp_elimination}
  \emph{proof subtyping (least fixed point elimination)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \J{LFP[}\alpha\J{]}\tau_l \subtypes \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \entails 
      \textbf{factor}(\alpha \entails \tau_l \cdot l) 
      \subtypes 
      \tau_r \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \J{LFP[}\alpha\J{]}\tau_l \subtypes \J{<} l \J{>} \tau_r 
      \given \Theta', \Delta' 
    }

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau \subtypes \tau_l \typdiff \tau_r \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \TODO{Update proof, deriving that lower does not intersect with subtraction}
      \\\\
      \TODO{in lean, prove that subtyping is equivalent to decidable definition under restricted subtyping form}
      \\\\
      \epsilon \pattype \tau_r
      \\
      \Theta, \Delta \restricted \tau_r \subtypes  \tau
      \\\\
      \Theta, \Delta \not\dentails \tau \subtypes \tau_r
      \\
      \Theta, \Delta \not\dentails \tau_r \subtypes \tau
      \\
      \Theta, \Delta \entails \tau \subtypes \tau_l \given \Theta', \Delta' 
    } {
      \Theta, \Delta \entails 
      \tau \subtypes \tau_l \typdiff \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\noindent
For least fixed point introduction \ms{
\tau_l \subtypes \J{LFP[} \alpha \J{]}\tau_r
}, the solver has a couple options.
One option is to check for a syntactic guarantee that the least fixed point is inflatable without diverging 
\ms{
... \circlearrowleft \tau_l \subtypes \J{LFP[} \alpha \J{]} \tau_r
} and then solve the inflated introduction.
Alternatively, it can strengthen the least fixed point
with union drop \ms{\textbf{drop}(\alpha entails \tau_r) = \tau_r'},
which drops any parts of the union containing the self reference.

@@

\noindent
Refinement elimination is the dual of expansion introduction, so its rules
rely on analogous techniques.
However, there is one additional rule
for eliminating an intersection and introducing an implication \ms{
\tau_l \& \tau_r \subtypes \tau_a \J{->} \tau_b
}.
As we mentioned earlier when discussing packing together the inferred types of 
multiple paths of functions, sometimes the argument type is weaker than each
path type antecedent, yet could be stronger than union of all the path type antecedents.    
As such, the solver may leverage path type merging \ms{
\textbf{merge}(\tau_l \J{\&} \tau_r)
} to weaken a function type
into a single path type where the antecedent and the consequent may consist
of compounded unions.

@@

\begin{requirement}
  \label{req:factorization_soundness}
  \emph{factorization soundness}  
  \small
  \begin{mathpar}
    \inferrule {
      \textbf{factor}(\alpha \entails \tau \cdot l) = \tau'
    } {
      \forall e' \qua
      e' \hastype \tau' 
      \iff 
      (\exists e \qua e \J{.} l \cong e' \up \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau)
    }
  \end{mathpar}
\end{requirement}
